\Title

### 2013-01-29 synmut.tex
edits in captions

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p8.pdf}
\label{fig:aftsyn}}\\
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p8.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of allele frequencies in \env, C2-V5, from
patient 8~\cite{shankarappa_consistent_1999},
divided by synonymous (panel A) and nonsynonymous (panel B) derived alleles.
While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e. its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}
\label{fig:fixp2}}
\caption{(Panel A) Time course of loss and fixation of synonymous mutations
observed in a frequency interval $\nu_0$. The ultimate fraction of synonymous
mutations that fix as a function of intermediate frequency $\nu_0$ is the
fixation probability. (Panel B) Fixation probability of derived synonymous
alleles is strongly suppressed in C2-V5 versus other parts of the {\it env}
gene. The horizontal error bars on the abscissa are bin sizes, the vertical ones
the standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
alleles (red line) are systematically larger than of alleles that are doomed to
extinction (blue line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to be
breaking RNA helices are also more likely to revert and finally be lost from the
population. As a control, the average over non-observed but potentially
available polymorphisms lies between the two curves (green line), as expected
(because only some of these mutations will interfere with stem loop formation).
To test the hypothesis that mutations in C2-V5 are lost since break stems
in the conserved between the variable loops, we split the synonymous mutations
in the extended V1-V5 region further into conserved and variable regions and
found that the biggest depression in fixation probability is observed in the
conserved stems, while the variable loops show little deviation from the
neutral signature, see \FIG{SHAPEB}. This is consistent with important stem
structures in conservered regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}
\label{fig:SHAPEA}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities. \comment{Is this C2-V5 only?} 
Panel A) Synonymous substitions in the C2-V5 region have substantially higher
SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost. This difference is quantified by a cumulative histogram
($p=0.002$, Kolmogorov-Smirnov two sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B shows the fixation probablility of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{(Panel A) The depression in $P_\text{fix}$ depends on the deleterious
effect size of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
(Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity. Parameters are chosen uniformly on a
logarithmic scale from the indicated windows. \comment{should we extend the
parameters space maybe to $10^{-4}$?}}
\label{fig:simheat}
\end{center}
\end{figure}

>>>>

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p8.pdf}
\label{fig:aftsyn}}\\
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p8.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of allele frequencies in \env, C2-V5, from
patient 8~\cite{shankarappa_consistent_1999},
divided by synonymous (panel A) and nonsynonymous (panel B) derived alleles.
While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations. Panel A shows the time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

@@

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
alleles (red line) are systematically larger than of alleles are lost (blue
line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. As a control, the
average over non-observed but potentially available polymorphisms lies between
the two curves (green line), as expected (because only some of these mutations
will interfere with stem loop formation).
To test the hypothesis that mutations in C2-V5 are lost since break stems in the
conserved between the variable loops, we split the synonymous mutations in the
extended V1-V5 region further into conserved and variable regions and found that
the biggest depression in fixation probability is observed in the conserved
stems, while the variable loops show little deviation from the neutral
signature, see \FIG{SHAPEB}. This is consistent with important stem structures
in conservered regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.48\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[width=0.48\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities.
Panel A) Synonymous substitions have
substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost (restricted to the vicinity of variable loops). This difference
is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B shows the fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity. Parameters are chosen uniformly on a
logarithmic scale from the indicated windows. This simulation assumes 6
roughly equivalent escape mutations per epitope; see supplement for alternative
models. }
\label{fig:simheat}
\end{center}
\end{figure}
